Delayed effects of climate on vital rates lead to demographic divergence in Amazonian forest fragments

### 2021-09-01 review/Response to reviews.Rmd

# Reviewer 1

**General comments**

> Scott and colleagues present the results of an interesting series of analyses designed to disentangle the lagged effects of climate and forest fragmentation on the demographic performance of a tropical understory plant.
> As ecologists seek to understand not just how changes in mean climate, but also changes in extremes, will influence population persistence, studies such as this that capitalize on long time series are critical to making progress.
> The authors find that the effects of swings in precipitation (both drought and particularly wet years) affect survival, growth and flowering, but that these effects are both lagged and differ by whether plants were in continuous forest or the smallest fragment size (1 ha).
> Overall, it's a well written paper, and an application of a relatively new modeling approach to address the questions.
> The results highlight both the challenges with understanding the complex factors driving population persistence and also how critical it will be to grapple with them.
> Below I provide suggestions for improvement, particularly aimed to make the new modeling approach more accessible.

We are delighted that the reviewer found the manuscript engaging and valuable.
Our responses to specific suggestions follow.

@@

##  {#R1-1}

> By the end of the introduction, I expected that the paper would focus on drought, and was then puzzled by the choice of the words 'extreme precipitation' -- this of course makes sense later that the results are focused both on the effects of drought and extreme wet.
> The introduction could set up the reader better to understand both.
> Since drought is well covered, adding background and hypotheses for why extreme wet would have a negative effect on vital rates would help.

We edited a sentence in the introduction that cites Esteban et al. (2021) to clarify that both extreme dry and wet conditions reduce growth and survival of trees in the Amazon.
*"...recent results suggest increases in the frequency and severity of extreme precipitation events **(both drought and extreme wet)** reduce survival and reproduction...".* Additionally, we added a sentence describing forecasted changes in precipitation in the Amazon.

@@

> One challenge of using a new modeling approach is making the methods of the paper understandable and to stand alone.
> As written in the methods and results, I found the implementation very difficult to follow.
> A slightly longer summary of the methods would help.

We have added an additional paragraph to the methods describing how the crossbasis smooth was fit using the `mgcv` and `dlnm` packages.
<!--# add line reference before submiting! --> We are not entirely sure if this is the type of description the reviewer was looking for.
If the reviewer(s) or editor see this as too detailed for the main text, we are happy to remove it, move it to a supplement, or simply replace it with a reference to the (now available) archived code for the analysis.
See also our response to \@ref(R3-3).

> In the results, I was surprised to see only a few p-values and not much else about the models in terms of how they were assessed for goodness of fit.
> While I am not at all suggesting that p-values are the solution, I would have liked a little more sense of how well the models fit.
> One way to do that might be to show the raw data plotted with the model fits so the reader can see, for example, what kind of functions were drawn.
> Another might be to present something in table format (the comparable version of a table of AICs or of the results of frequentist tests).

Rather than comparing the fit of multiple models with different predictors, we took advantage of the `select = TRUE` option in the `gam()` function, which adds an additional penalty to smooth terms that effectively removes them from the model if they are not supported by data.
Therefore we cannot show something like a dAIC table.
However, we have now included a table with R^2^ values for each model indicating fit as well as estimated degrees of freedom and p-values for each term (Table 2).
We have replaced many of the parenthetical p-values reported in the results section with references to this table.

@@

##  {#R1-4}

> Related to above, I think it's important to see more of the raw data than are presented in Fig.
> 1.
> (I'm thinking of a version of Fig 3 that shows size in t+1, survival, and flowering rather than effect size on the y-axis, with the data and lines showing the shape of the GAMs).
> This could be an appendix, but it would help with interpreting Figures 3 -- 6.

We changed Fig 3 to be on the response scale as suggested (see our response to \@ref(R1-figs)), but showing the raw data made the plots visually cluttered, in our opinion.
Another way we approached this comment is by excluding areas of the heatmaps in figures 4-6 that are far from observed data, equivalent to not drawing regression lines beyond the range of data.
We implemented this with the `exclude.too.far()` function in the `mgcv` package.
Displaying raw data on the heatmaps was too messy and not informative in a 2d plot.

@@

> I suggest that a little more about the limitations of this modeling approach in the discussion would help set the context for interpreting the results and make the paper more robust.

The main limitations of this method are of data (see also reviewer 2's comment \@ref(R2-15)).
We have added some text to the discussion describing limitations found in previous simulation studies, and new unpublished evidence that spatial replication may be more valuable than long time series for detecting demographic effects of the environment.

> In some ways it seems like the analysis is adding more precision (by using monthly climate data) than the plants perceive and that would be measured in a yearly census.

Counter-intuitively, averaging weather conditions over an entire year may obscure ways plants perceive environmental variation.
For example, a drought in the wet season may have a different effect than a drought in the dry season.
This differential response to season- or month-level climate data is observable in yearly plant census.
Collapsing finer temporal scale environmental drivers to yearly averages (or a few means of arbitrarily chosen climate windows) is exactly what we avoid with this method.
The caveat to this is that without finer scale demographic data, one may not be able to discern the cause of an apparent delayed effect, which is why we expand on several possible mechanisms for delayed effects in detail in the discussion.

@@

Our point was actually that SPEI had less of an impact on flowering probability in fragments.
We edited this sentence to read: "SPEI had less of an effect on flowering probability in 1-ha fragments as indicated by the narrower range of the evaluated smooth (Figure 6b)."

##  {#R1-8}

> L442-443 -- What is a mechanism for why plants might flower more after drought?
> (the discussion talks about the opposite) The next paragraph kind of goes there, but the link could be tighter.

We are a little hesitant to speculate too much here, as not much is known about the optimal microclimate conditions for flowering in *H. acuminata*, but we have revised this paragraph (see also our response to \@ref(R2-17)).
It now includes the following possible explanation: *"Drought conditions could be favorable for H. accuminata flowering due to increased temperatures or decreased cloudiness associated with droughts (Pau et al. 2013),* *and the effects could be delayed due to the development time of inflorescences."*

@@

> Fig 2 -- This was only cited in the discussion as far as I can tell.
> Bring it back into the results?

We have added a sentence about this figure in the results.

> Fig 3A -- The results mention something about survival approaching 1 -- I wasn't sure if that was supposed to be visible here.

This was referring to a previous iteration of the figure where the y-axis was on the response scale.
We have reverted the figure to this version (see response below) and this should be more obvious now.

@@

# Reviewer 2

**General comments**

> The manuscript by Scott et al analyzed an eleven year long demographic data set on the herb, Heliconia acuminata in the Central Amazon to evaluate the effect of drought and fragmentation on survival growth and fecundity.
> The authors also employ distributed lag non-linear models to explore delayed effects of drought.
> This work is an important contribution adding to the sparce literature documenting the role of weather variables on plant vital rates.
> The manuscript is generally well written.
> I hope the authors will find my comments helpful in improving this already very good manuscript.

We are glad to hear the reviewer saw this work as important and well written.
We address specific concerns and suggestions below.

@@

We have edited the abstract.
Now the sentence to which the reviewer refers reads: *"We used distributed lag non-linear models to understand the delayed effects of standardized precipitation evapotranspiration index (SPEI) on survival, growth, and flowering."*

##  {#R2-2}

> Line 51-52.
> I find this sentence confusing.
> What specific demographic mechanisms are you referring to?
> Do you mean that high mortality rate lead to extinction and we don't know what causes high mortality rates?

The reference, Bruna et al. (2009), discusses the lack of knowledge of what demographic vital rates (e.g. survival, reproduction, recruitment) in which life stages (e.g. small or large individuals) contribute most to population declines in fragments.
We have re-ordered this sentence to emphasize the hypothesis that fragmentation alters survival, growth, or reproduction with the caveat that few studies have determined which of these vital rate changes are driving extinctions.

@@

This is just the height of the plant from the ground to the tallest leaf leaf tip, not the tip of the youngest leaf.
The reviewer is correct that the height of a plant growing more horizontally (for example if it was knocked over by a branchfall) could be shorter than the length from the base to the end of the tallest leaf.
We edited the description to read: "...the height of each plant (i.e. distance from the ground to the tallest leaf tip)"

##  {#R2-4}

> Lines 251ff: I don't see a reference of Fig 2 in the result section.

We have added a sentence to the results that refers to this figure.

@@

We have made this correction.

##  {#R2-17}

> Line 440: Can you explain a bit more in detail why drought increasing flowering probability two years in the future would be a bet hedging response?

We intended this paragraph to focus on one possible explanation for delayed effects---delays due to development times.
We believe that the bet-hedging argument was confusing for two reasons: 1) it assumed that drought (as we measured it with SPEI) is detrimental to *H. accuminata* survival, which is not consistently reflected in our results, and 2) it did not address longer lags than the development time of a flowering shoot.
We have therefore revised the example in this paragraph to focus on literature that shows positive effects of temperature and negative effects of cloudiness on flowering in tropical plants---weather conditions both associated with droughts.
See also our response to \@ref(R1-8).

@@

##  {#R2-18}

> Line 448: Alternatively, drought may reduce nutrient availability such that after covering maintenance costs there is nothing left to allocate.

We thank the reviewer for the suggestion---in fact, effects of drought on maintenance costs is precisely what we were trying to communicate in this line.
After shedding shoots and leaves during drought, maintenance costs may be reduced, and plants might invest more in rhizomes, possibly allowing plants to recover after drought to a larger size.
We are unaware of literature investigating the belowground effects of drought on *Heliconia*, but this is consistent with results in Bruna et al. 2002 (see response to \@ref(R2-19)).

@@

##  {#R2-19}

> Lines 451-453: I think herbivory is this is very different from drought.
> Drought reduced nutrient availability while herbivory removes some plant biomass triggering compensatory growth.

We agree with the reviewer and have replaced this sentence with a more in-depth explanation of Bruna et al. 2002.

@@

We agree with the reviewer that the exact timing of mortality within a season doesn't matter for population dynamics.
This is likely obvious to demographers who are used to working with yearly census data and finer scale climate data.
However, we think this paragraph is appropriate to follow up with the three before it---a delayed effect indicated by a DLNM plot like those in Figures 4, 5 and 6 could be due to 1) developmental delays, 2) buffering by below-ground storage tissues, 3) indirect effects, or 4) **none of the above.** We want to be sure to point out that in the case of survival especially, a "delayed effect" of SPEI 5 months in the past could simply mean that a plant died *during* a drought 5 months in the past (no actual delay between adverse conditions and death).
Importantly, this does not apply to flowering in our system, since flowering happens at a specific time of year (during the census), unlike mortality.
We have revised this paragraph to emphasize the above points.

##  {#R2-21}

> The discussion includes reasons for why we expect delayed effect in plants in general, but it does not relate the specifics of the significant delays to the life history of H.
> acuminata.
> For instance, why does too much rain reduce survival?
> Why is there an effect of SPEI during the dry season (18-20 months in the past but during the wet season 32-36 months in the past? Why do plants in fragments have a decrease survival when precipitation is high during dry and wet periods (1-18 months time lag)?
> Etc.

While we believe we did discuss *Heliconia*-specific mechanisms for delayed responses of flowering (L 434--438, now <!--# new line -->) and delayed effects on growth (L 445--450, now <!--# new line -->), the reviewer is correct that we had not previously discussed mechanisms behind mortality or why wet conditions might be detrimental.
We have added a paragraph to the discussion regarding effects of wet conditions on tropical plants.

@@

Some of the missing values in the data do represent *Heliconia* plants that lost all above-ground tissue but re-sprouted in successive years, but other missing values may simply be imperfect "recapture".

##  {#R3-2}

> Line 166 - given the shallow roots, why not use 1 month SPEI and let the model do the smoothing over longer time frames?
> You describe why not using longer smooths, but not shorter.

\
The choice of scale for the SPEI calculation is indeed difficult, and is related to the operational definition of drought (i.e. is drought defined as dry conditions over the previous 1, 2, 3 or more months?).
We are not certain that the reviewer's suggestion to allow the GAM to "do the smoothing" would have the same effect as changing the scale parameter when calculating SPEI.
SPEI is calculated using a sliding window, while GAMs use penalization to achieve smoothing.
We believe a thorough comparison of the ability for different SPEI scales to explain variation in survival, size, and flowering is beyond the scope of this paper.

### 2021-09-01 doc/paper.Rmd
fixed table captions by switching to pander.  Closes #97

```{r setup, include=FALSE}
knitr::opts_chunk$set(
	echo = FALSE,
	fig.path = "figures/",
	dev = "png",
	dpi = 300,
	message = FALSE,
	warning = FALSE,
	collapse = TRUE,
	comment = "#>"
)
library(targets)
library(tidyverse)
library(tsibble)
library(lubridate)
library(knitr)
library(here)
library(patchwork)
```

@@

@@

Demographic responses will also be delayed if abiotic stress causes plants to invest in belowground rhizomes [*sensu* @pumisutapon2012].
The carbohydrates stored in rhizomes allow *Heliconia* to regenerate aboveground biomass following damage [@rundelStructuralPhysiologicalAdaptation1998] and protect the buds that give rise to new shoots from stressful conditions [@klimesovaBelowgroundPlantFunctional2018].
This may be why drought led to delayed increases in growth---by shedding shoots and leaves [@brunaExperimentalAssessmentHeliconia2002] and investing in rhizomes, plants are generating proportionately more buds with which to regenerate when conditions improve.
This would also be consistent with the results of prior experiments, in which the growth rates of *H. acuminata* 8 months after they were mechanically damaged far exceeded those of control plants [@brunaCompensatoryResponsesUnderstory2005].

@@

Finally, demographic delays could be be demographic only, rather than due to physiological delays or indirect effects.
For example, if a severe month-long drought 6 months prior to a census caused increased mortality, it would appear in our models as a delayed effect, even if plants died immediately within that month, because mortality was only captured once per year.
In our case, this potential explanation for delayed effects applies only to plant size and survival, as the flowering season coincided with the yearly census.
It also only applies to lags of up to 12 months since mortality, size, and flowering are recorded yearly.
This possibility is not unique to our study, rather it is a consequence of conducting demographic censuses on an annual scale while the climate is quantified monthly or seasonally.
While the very slow growth and extremely low mortality rates of *H. acuminata* mean this effect is unlikely to be acting in our system, it may be that for some species it will be important to conduct demographic surveys at the same temporal scale at which climate is aggregated to better understand delayed effects of climate.
<!--# might need some other edits here to address why it doesn't matter for demography which of these explanations is true.  Might be a better sentence than the previous sentence -->

@@

Finally, any analytical approach for assessing lagged effects on demography requires long-term data and/or data collected from many sites with independent weather (but similar climate and site conditions) [@tenhumbergTimelaggedEffectsWeather2018; @eversLaggedDormantSeason2021]. @tellerLinkingDemographyDrivers2016 used a simulation study to show that detecting lagged effects required 20--25 years of data, although this depended on effect size and the range of climate extremes in the data. Compagnoni et al. ( <!--# cite ms in prep?  cite ESA talk? --> ) showed that spatial replication can alleviate this requirement, and in fact in our study we were able to detects significant lagged effects with only 10 years of data, but multiple plots experiencing slightly different weather. Unfortunately, long-term data monitoring the entire life-cycle of tropical taxa are rare, and those doing so in fragmented landscapes are virtually nonexistent [@brunaCompensatoryResponsesUnderstory2005].
Without investing in collecting such data, generalizations regarding the demographic consequences of climate change in these species rich and increasingly fragmented habitats will continue to prove elusive.
More generally, however, researchers need to consider how delayed responses to climate could influence the interpretation of data in studies where the organisms lifespan exceeds the study's duration.

@@

We thank Collin Edwards, Andrew Mercadante, Ellie McDaniel, \_\_\_, \_\_\_, and \_\_\_ anonymous reviewers for helpful discussions and comments on the manuscript.
We also thank the technicians and students who helped conduct the *Heliconia* censuses and the BDFFP and INPA for logistical support.
Financial support was provided by the U.S.
National Science Foundation (awards DEB-0614339, DEB-0614149, INT 98-06351, and DEB-1948607).
This is publication No. \_\_\_ in the BDFFP Technical Series.
The authors delcare no conflicts of interest.

@@

```{r}
withr::with_dir(here::here(), {
  targets::tar_load(c(
    intercept_df,
    results_df
  ))
})
flextable::set_flextable_defaults(font.family = "Cambria", font.size = 10)
```

```{r}
ci_cap <- "Model intercepts and their 84% confidence intervals.  Non-overlapping confidence intervals are bolded and can be interpreted as a significant difference in the vital rate between continuous forest and 1 ha fragments."
```

```{r label="intercept-table"}
flextable(intercept_df) %>% 
  autofit() %>% 
  merge_v(j = 1) %>% 
  valign(j = 1, valign = "top") %>% 
  align(j = 3, align = "right", part = "all") %>% 
  bold(j = 3, i = c(3:6)) %>% 
  fix_border_issues() %>%
  set_caption(ci_cap, style = "Image Caption")
```

@@

```{r}
result_cap <- 'Marginal hypothesis tests for models of survival probability, size, and flowering probability of *H. acuminata* in continuous forest and 1 ha fragment plots. The adjusted R^2^ is reported as a measure of model fit. The model terms included the parametric fixed effect factor of whether plants flowered the previous year ("flowered"), the smoothed fixed effect of plant size ("s(log(size~t~))"), the crossbasis smooth of lagged SPEI ("s(SPEI, lag)"), and a random effect of plot ID, ("s(plot)"). Degrees of freedom are reported for the parametric term, and estimated degrees of freedom (edf) are reproted for smooths.  Larger values for edf indicate more complex smooths and when edf is zero the term is effectively dropped from the model. The test statistic reported is $\\chi^2$ for survival and flowering and $F$ for log(size~t+1~).'
```

<!--# add something about how p-values are appropriate for random effects?  see help files from mgcv for references -->

```{r label="results-table"}
flextable(results_df) %>% 
  merge_v(j = c(1,2,3)) %>% 
  valign(j = c(1,2,3), valign = "top") %>% 
  align(j = c(5, 7), align = "right", part = "all") %>%
  colformat_double(j = c(3,6), digits = 2) %>% 
  fix_border_issues() %>% 
  flextable::compose(j = 3, part = "header",
                     value = as_paragraph("model R", as_sup("2"))) %>% 
  flextable::compose(j = 4, i = c(2,6,10,14,18,22),
                     value = as_paragraph("s(log(size", as_sub("t"), "))")) %>%
  flextable::compose(j = 1, i = 9, 
                     value = as_paragraph("log(size", as_sub("t+1"), ")")) %>% 
  autofit() %>% 
  set_caption(result_cap, style = "Image Caption")
```

>>>>

```{r setup, include=FALSE}
knitr::opts_chunk$set(
	echo = FALSE,
	fig.path = "figures/",
	dev = "png",
	dpi = 300,
	message = FALSE,
	warning = FALSE,
	collapse = TRUE,
	comment = "#>"
)
library(targets)
library(tidyverse)
library(tsibble)
library(lubridate)
library(knitr)
library(here)
library(patchwork)
library(pander)
```

@@

Interestingly, we found significant detrimental effects of unusually wet conditions at some lag times for all vital rates except survival in continuous forest and flowering probability in 1 ha fragments.
Although the literature focuses primarily on the effects of droughts in the tropics, some research shows that unusually wet conditions can also be detrimental to tropical plants [@estebanOtherSideDroughts2021].
Wet conditions could be associated with increased cloud cover which could reduce photosynthetic activity and decrease growth, or with wind storms which could increase tree- and branch-fall---a source of mortality for understory plants.
Waterlogged soils can also decrease growth and increase mortality [@parentOverviewPlantResponses2008].
Finally, cool temperatures associated with unusually wet conditions may decrease flowering [@pauCloudsTemperatureDrive2013].

@@

Demographic responses will also be delayed if abiotic stress causes plants to invest in belowground rhizomes [*sensu* @pumisutapon2012].
The carbohydrates stored in rhizomes allow *Heliconia* to regenerate aboveground biomass following damage [@rundelStructuralPhysiologicalAdaptation1998] and protect the buds that give rise to new shoots from stressful conditions [@klimesovaBelowgroundPlantFunctional2018].
This may be why drought led to delayed increases in growth---by shedding shoots and leaves and investing in rhizomes, plants could be generating proportionately more buds with which to regenerate when conditions improve.
This is consistent with the results of prior experiments in which *H. acuminata* plants transplanted into hotter, drier fragments lost and then recovered far more leaf area than control plants[@brunaExperimentalAssessmentHeliconia2002].

@@

Finally it is important to clarify the distinction between "physiological delays" and "observational delays".
For example, a Fall drought could reduce cold tolerance and therefore overwinter survival.
Alternatively the drought could kill plants immediately.
In both cases, if mortality is recorded in the Spring, both would appear as a delayed effect of the drought, but only the first case is a truly delayed physiological response to drought, the apparent delayed response in the second case is due to the timing of the response in relation to the census date.
In our case, this potential explanation of "observational delays" applies only to plant size and survival, as the flowering season coincided with the yearly census.
It also only applies to lags of up to 12 months since mortality, size, and flowering are recorded yearly.
This possibility is not unique to our study, rather it is a consequence of conducting demographic censuses on an annual scale while the climate is quantified monthly or seasonally.
For some demographic models and follow-up experiments, knowing the precise timing of mortality might be critical; for others less so.
To disentangle possible mechanisms for observed delayed effects, it may be necessary to conduct demographic surveys at the same temporal scale at which climate is aggregated.

@@

Finally, any analytical approach for assessing lagged effects on demography requires long-term data and/or data collected from many sites with independent weather (but similar climate and site conditions) [@tenhumbergTimelaggedEffectsWeather2018; @eversLaggedDormantSeason2021]. @tellerLinkingDemographyDrivers2016 used a simulation study to show that detecting lagged effects required 20--25 years of data, although this depended on effect size and the range of climate extremes in the data. @compagnoni2021 showed that spatial replication can alleviate this requirement, and in fact in our study we were able to detects significant lagged effects with only 10 years of data, but multiple plots experiencing slightly different weather. Unfortunately, long-term data monitoring the entire life-cycle of tropical taxa are rare, and those doing so in fragmented landscapes are virtually nonexistent [@brunaCompensatoryResponsesUnderstory2005].
Without investing in collecting such data, generalizations regarding the demographic consequences of climate change in these species rich and increasingly fragmented habitats will continue to prove elusive.
More generally, however, researchers need to consider how delayed responses to climate could influence the interpretation of data in studies where the organisms lifespan exceeds the study's duration.

@@

We thank Collin Edwards, Andrew Mercadante, Ellie McDaniel and 3 anonymous reviewers for helpful discussions and comments on the manuscript.
We also thank the technicians and students who helped conduct the *Heliconia* censuses and the BDFFP and INPA for logistical support.
Financial support was provided by the U.S.
National Science Foundation (awards DEB-0614339, DEB-0614149, INT 98-06351, and DEB-1948607).
This is publication No. \_\_\_ in the BDFFP Technical Series.
The authors delcare no conflicts of interest.

@@

```{r}
withr::with_dir(here::here(), {
  targets::tar_load(c(
    intercept_df,
    results_df
  ))
})
```

```{r intercept-table, results='asis'}
intercept_df[c(2,4,6), 1] <- ""
emphasize.strong.cells(as.matrix(data.frame(row = 3:6, col = rep(3,4))))
pandoc.table(
  intercept_df,
  justify = c("left", "left", "right"),
  full_width = FALSE,
  caption = "(\\#tab:intercept-table) Model intercepts and their 84% confidence intervals.  Non-overlapping confidence intervals are bolded and can be interpreted as a significant difference in the vital rate between continuous forest and 1 ha fragments."
)
```

@@

```{r results-table, results='asis'}
results_tab <- 
  results_df %>% 
  mutate(across(c(R2, `test statistic`),
                ~round(.x, 2) %>% 
                  prettyNum(big.mark = ","))) %>% 
  mutate(term = case_when(
    term == "s(log(size_t))" ~ "s(log(size~t~))",
    term == "flowered" ~ "flowering~t~",
    TRUE ~ term
  ),
  `vital rate` = case_when(
    `vital rate` == "size_next" ~ "log(size~t+1~)", 
    `vital rate` == "flowering" ~ "flowering~t+1~",
    TRUE ~ `vital rate`)) %>% 
  rename("R^2^" = R2)
results_tab[c(2:8, 10:16, 18:24), 1] <- ""
results_tab[c(2:4, 6:8, 10:12, 14:16, 18:20, 22:24), 2:3] <- ""

pandoc.table(
  results_tab,
  full_width = FALSE,
  split.table = 100,
  justify = c("left", "left", "right", "left", "right", "right", "right"),
  caption = '(\\#tab:results-table) Marginal hypothesis tests for models of survival probability, size, and flowering probability of *H. acuminata* in continuous forest and 1 ha fragment plots. The adjusted R^2^ is reported as a measure of model fit. The model terms included the parametric fixed effect factor of whether plants flowered the previous year ("flowering~t~"), the smoothed fixed effect of plant size ("s(log(size~t~))"), the crossbasis smooth of lagged SPEI ("s(SPEI, lag)"), and a random effect of plot ID, ("s(plot)"). Degrees of freedom are reported for the parametric term, and estimated degrees of freedom (edf) are reproted for smooths.  Larger values for edf indicate more complex smooths and when edf is zero the term is effectively dropped from the model. The test statistic reported is $\\chi^2$ for survival and flowering and $F$ for log(size~t+1~).'
)
```
